---
title: "tibblify"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{tibblify}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r packages-used}
library(tibblify)
library(purrr)
library(vctrs)
```

## Introduction

With `tibblify()` you can rectangle deeply nested lists into a tidy tibble. 
These lists might come from an API in the form of JSON or from scraping XML.

## Example

Let's start with `gh_users`, which is a list containing information about four GitHub users. 
We'll select a subset of columns to keep the example relatively simple.

```{r}
library(tibblify)

gh_users_small <- purrr::map(gh_users, ~ .x[c("followers", "login", "url", "name", "location", "email", "public_gists")])

names(gh_users_small[[1]])
```

We can quickly rectangle `gh_users_small` with `tibblify()`.

```{r}
tibblify(gh_users_small)
```

We can now look at the specification `tibblify()` used for rectangling.

```{r}
guess_tspec(gh_users_small)
```

If we are only interested in some of the fields we can easily adapt the specification.

```{r}
spec <- tspec_df(
  login_name = tib_chr("login"),
  tib_chr("name"),
  tib_int("public_gists")
)

tibblify(gh_users_small, spec)
```


## Objects

We refer to lists like `gh_users_small` as _collections_, and each element of that list is an _object_. 
Objects and collections are the typical input for `tibblify()`.

Basically, an _object_ is something that can be converted to a one-row tibble. 
This boils down to a condition on the names of the object:

1. The `object` must have names (the `names` attribute must not be `NULL`).
2. Every element must be named (no name can be `NA` or `""`).
3. The names must be unique.

In other words, the names must fulfill `vctrs::vec_as_names(names, repair = "check_unique")`. 
The name-value pairs of an object are the _fields_.

For example `list(x = 1, y = "a")` is an object with the fields `(x, 1)` and `(y, "a")` but `list(1, z = 3)` is not an object because it is not fully named.

A _collection_ is a list of similar objects so that the fields can become the columns in a tibble.

## Specification

Providing an explicit specification has several advantages:

1. You can ensure type and shape stability of the resulting tibble in automated scripts.
2. You can give the columns names that are different than their names in the source collection.
3. You can parse only the fields you need (and ignore any others).
4. You can specify what happens if a value is missing.

To create a specification for a collection, use `tspec_df()`. Describe the columns of the output tibble with the `tib_*()` functions. The `tib_*()` functions describe the path to the field to extract and the output type of the field. Throughout these functions, we use the concept of `ptype` (prototype) as used in the {vctrs} package (see `vignette("type-size", package = "vctrs")`).

There are five main types of `tib_*()` functions:

* `tib_scalar(ptype)`: A length-one vector with type `ptype`. The result is a column of that type.
* `tib_vector(ptype)`: A vector of arbitrary length with type `ptype`. The result is a list column, with each row containing a vector of that type.
* `tib_variant()`: A vector of arbitrary length and type. The result is a list column, with each row containing a list. You should rarely need this function.
* `tib_row(...)`: An object with the fields `...`. The result is a tibble column, with each row containing the specified one-row tibble.
* `tib_df(...)`: A collection where the objects have the fields `...`. The result is a tibble column, with each row containing a tibble.

There are also two other `tib_*()` functions for special cases:

* `tib_recursive()`: A collection where objects can themselves be collections of the same structure, such as a directory tree. The result is a tibble column, with each row containing a tibble of the same structure (until `NULL` is reached, terminating recursion).
* `tib_unspecified()`: An unspecified field where the type and shape is not known in advance. The `unspecified` argument of `tibblify()` controls how such fields are handled.

For convenience there are shortcuts for `tib_scalar()` and `tib_vector()` for
the most common prototypes:

* `logical()`: `tib_lgl()` and `tib_lgl_vec()`
* `integer()`: `tib_int()` and `tib_int_vec()`
* `double()`: `tib_dbl()` and `tib_dbl_vec()`
* `character()`: `tib_chr()` and `tib_chr_vec()`
* `Date`: `tib_date()` and `tib_date_vec()`
* `Date` encoded as character: `tib_chr_date()` and `tib_chr_date_vec()`

### Scalar Elements

Scalar elements are the most common case and result in a normal vector column.

```{r}
tibblify(
  list(
    list(id = 1, name = "Peter"),
    list(id = 2, name = "Lilly")
  ),
  tspec_df(
    tib_int("id"),
    tib_chr("name")
  )
)
```

With `tib_scalar()`, you can also provide your own prototype for column types not included in our shortcuts. 
For example, let's say you have a list with durations (objects with class "difftime").

```{r}
x <- list(
  list(id = 1, duration = vctrs::new_duration(100)),
  list(id = 2, duration = vctrs::new_duration(200))
)
x
```

Use `tib_scalar()` with `vctrs::new_duration()` to specify the duration `ptype`.

```{r}
tibblify(
  x,
  tspec_df(
    tib_int("id"),
    tib_scalar("duration", ptype = vctrs::new_duration())
  )
)
```


### Vector Elements

If an element does not always have size one then it is a vector element. 
If it still always has the same type `ptype` then it produces a list column with elements of `ptype`.

```{r}
x <- list(
  list(id = 1, children = c("Peter", "Lilly")),
  list(id = 2, children = "James"),
  list(id = 3, children = c("Emma", "Noah", "Charlotte"))
)

tibblify(
  x,
  tspec_df(
    tib_int("id"),
    tib_chr_vec("children")
  )
)
```

You can use [`tidyr::unnest()`](https://tidyr.tidyverse.org/reference/nest.html) or [`tidyr::unnest_longer()`](https://tidyr.tidyverse.org/reference/hoist.html) to flatten these columns to regular columns.


### Object Elements

In `gh_repos_small`, the field `owner` is an object itself.

```{r}
gh_repos_small <- purrr::map(gh_repos, ~ .x[c("id", "name", "owner")])
gh_repos_small <- purrr::map(
  gh_repos_small,
  function(repo) {
    repo$owner <- repo$owner[c("login", "id", "url")]
    repo
  }
)

gh_repos_small[[1]]
```

The specification to extract it uses `tib_row()`.

```{r}
spec <- guess_tspec(gh_repos_small)
spec
```

This specification results in a tibble column.

```{r}
tibblify(gh_repos_small, spec)
```

If you don't like the tibble column you can unpack it with `tidyr::unpack()`. 
Alternatively, if you only want to extract some of the fields in `owner` you can use a character vector path.

```{r}
spec2 <- tspec_df(
  id = tib_int("id"),
  name = tib_chr("name"),
  owner_id = tib_int(c("owner", "id")), # "id" in "owner"
  owner_login = tib_chr(c("owner", "login")) # "login" in "owner"
)
spec2

tibblify(gh_repos_small, spec2)
```


## Required and Optional Fields

Objects usually have some fields that always exist and some that are optional. 
By default `tib_*()` requires that a field exists.

```{r error=TRUE}
x <- list(
  list(x = 1, y = "a"),
  list(x = 2)
)

spec <- tspec_df(
  x = tib_int("x"),
  y = tib_chr("y")
)

tibblify(x, spec)
```

You can mark a field as optional with the argument `required = FALSE`.

```{r}
spec <- tspec_df(
  x = tib_int("x"),
  y = tib_chr("y", required = FALSE)
)

tibblify(x, spec)
```

You can specify the value to use with the `fill` argument.

```{r}
spec <- tspec_df(
  x = tib_int("x"),
  y = tib_chr("y", required = FALSE, fill = "missing")
)

tibblify(x, spec)
```


## Converting a Single Object

To rectangle a single object you have two options: `tspec_object()` which produces a list, or `tspec_row()` which produces a tibble with one row.

It often makes more sense to convert such objects to a list. 
For example a typical API response might be something like this.

```{r}
api_output <- list(
  status = "success",
  requested_at = "2021-10-26 09:17:12",
  data = list(
    list(x = 1),
    list(x = 2)
  )
)
```

Use `tspec_row()` to convert to a one-row tibble.

```{r}
row_spec <- tspec_row(
  status = tib_chr("status"),
  data = tib_df(
    "data",
    x = tib_int("x")
  )
)

api_output_df <- tibblify(api_output, row_spec)
api_output_df
```

To access `data` one has to use `api_output_df$data[[1]]`, which is not very nice. 
We can use `tspec_object()` to simplify this situation.

```{r}
object_spec <- tspec_object(
  status = tib_chr("status"),
  data = tib_df(
    "data",
    x = tib_int("x")
  )
)

api_output_list <- tibblify(api_output, object_spec)
api_output_list
```

Now accessing `data` does not require an extra subsetting step.

```{r}
api_output_list$data # No [[1]] needed
```
